Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2019-08-14 Response_to_Reviewers.tex

%Resubmit 15 Aug. 2019\\

\bigskip

We want to thank the two reviewers for their thorough evaluation and discussion of our paper, as well as the challenging questions they asked. We have revised the manuscript in a way that we believe addresses most of the comments. In this response, we first want to address the two main questions that arose for both reviews. The first one (A) relates to the original mis-specification/ambiguities between the social and spatial components of the modelling process. The second one (B) relates to the contribution of the approach in terms of potential use for model makers, for domain experts and in operational contexts.

 \medskip
A. In terms of the ambiguity between the social and spatial component 
\begin{itemize}
\item A.1. We have precised and clarified the argument of the paper (mainly in introduction) by qualifying our contribution, which is to provide an operational framework for studying the influence of geometric structures of space on the results of a geosimulation model. We have thus distinguished two effects of geographical space in simulation modelling: the ones that are part of the model, reflecting for example the role played by distance on decreasing agents' perception or potential social interactions; and the ones which act as geometric structures "on which" the social elements simulated interact, i.e. geometric constraints of the simulation. The first type of effects is addressed by traditional modelling. Our original contribution is to tackle the second type.

\item A.2. We have clarified the terms used in our paper introduction to distinguish between: \textbf{sociospatial systems} (i.e. groups of social agents whose behaviour is constrained by their position in geographical space), \textbf{interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and \textbf{geometric structure}, a term we use for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model. When we talk about the generator we actually used in the paper, we call it a "density grid generator" because it generates a square grid with heterogeneous density, whereas in all generality a "spatial generator" could potentially generate graph or point structures.

@@

 \item A.3. We have modified figure 1 to express the clarified approach visually. Naming the generator parameter "gamma" has helped reduced the confusion that the term 'meta-parameter' was inducing. We have also replaced the expressions "quantitative results" by "sensitivity analysis" and "qualitative results" by "variations by types of geometric structure", in the figure as in the subsections of the paper.
\end{itemize}

 \medskip
B. To highlight limitations as well as promising perspectives of improvement, we answer the concern about the genericity of the method and expose its links with the conceptual and empirical domains:

 \begin{itemize}

  \item B.1. We have added a discussion of on how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities). This is a limitation of the method as it is but also an avenue for promoting stronger indicators in empirical research.

  \item B.2. To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, we added a paragraph in discussion, explaining how our approach allows us to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences).

  \item B.3. As a perspective to reduce this bias, we could compute ego-neighbourhoods rather than a regular grid, and make these systematically vary and study the effect on emergent behavior and indicators. In further researches, we plan to extend the analysis to vary the size of grid and its cell shape (cf. MAUP), but also generate other types of geometric structures to feed into geosimulation models (for example points or networks).
\end{itemize}

@@

  \item \textit{Then, in paragraph 1.3, the complexity of systems with respect to segregation is also problematically described.} \ldots
  % Emergent behaviour that may lead to segregation is first and foremost a social process, taking social interactions at the local scale (income, housing costs, etc.) and/or at the macro scale (laws, rules, economic system, cultural attitudes, etc.) into account. These interactions can then be associated with the material and relational dimension of geographical spaces. The question remains: what is the role of geographical systems? In this respect, the paper offers some relevant clues such as the MAUP problem or the “uncertain geographic context problem”.

  $\rightarrow$ see comment A. As for the role of geometric structures in the production of segregation, Banos (2012) has shown that with the same Schelling mechanisms and the same parameter values but different network structures of neighbourhoods (scale-free, regular, etC.) the segregation observed at the end of simulation varied significantly between network structures. This is the kind of approach we systematise here, although with a spatial grid instead of network structures.

  \medskip

  \item \textit{The problem of setting space and geography as synonymous continues in paragraphs 1.9 and following. What is described here as "spatial system shape" is the geometry of spaces but not the geography of spaces.}

@@

  \item \textit{Due to this, attempts to explain the growth of existing cities should be avoided.}

  $\rightarrow$ We agree with this comment and recall that urban growth processes are not the main focus of the paper, they are here an instantiation of a given $\gamma$ generator, which could have been replaced by others such as network models, procedural modeling, etc. The position and discussion of the paper was amended to precise that (see comment A).

  \medskip

  \item \textit{The approach is reductionist and has its strength in reflecting growth at an abstract level nevertheless. The concluding remark in paragraph 5.1 “With the Schelling experiment, we found … that polycentric and discontinuous cities appear systematically more segregated than compact cities …” is elusive. Either it is a trivial fact, simply because polycentric cities have more centres and thus, by nature, more opportunities from which segregation can originate. Or, the results can be used as a starting point to dig deeper in geographical reasons, taking social, economic, cultural, and political, but also topographical issues into consideration. This geographical reasoning, however, is outside of the scope of this paper. }

  $\rightarrow$ We understand this statement, however we wish to underline that our objective in the discussion of the Schelling results was not to establish a general result on the links between polycentricity and segregation, which are indeed very complex and multidimensional as advocated by the reviewer. Instead, we aimed at showing what kind of results could be obtained using the framework we suggest. If we do not prove that in general, and in empirical context, polycentricity creates more segregation, we obtained, in the modelling context, under Schelling assumptions, that polycentric geometric structures are more often that not associated with segregated patterns, a result which we do not believe to be solely resulting from the fact that more centers leads to more opportunities of segregation (which is true), because it seems also true that monocentric structures, with higher densities, can more easily accommodate heterogeneity of agents within cells, and hence a more patchy type of segregation which can for instance being observed in London. We included a comment in the discussion section, saying that we hope we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results.

 % \comment{in discussion, say that indeed the indicators can themselves be influenced by geometric structures (cf. comment B) but on the processes, there is nothing obvious about the influence of structures on behavioural spatial patterns (if you imagine a model with no interactions between agents for example) + it's a difference between the model and its potential addition to the real systems. cf. B}

   \item \textit{Formal issues that have to be solved: the numbering of paragraphs is not coherent. The reference of Wheeler (2006) is missing the source of the article.}

   $\rightarrow$ These issues were corrected.
